Do syntactic and semantic similarity lead to interference effects?\\ Evidence from self-paced reading and event-related potentials using German

### 2024-11-11 paper/revision2/ResponseToReviewers.tex

I do have some suggestions. In my view these are only minor. They mainly concern just the way things are presented or discussed. I list my comments by page.

p. 16: osf data etc.

I wanted to check those, but \textbf{when I clicked the link, it said I don't have permissions and that I need to request access}.''
\end{quote}

\subsubsection*{Response to R1.1}
A formatting error broke the link in the previous version of the manuscript. We have made the OSF repository public and updated the link in the manuscript. This should grant general accessibility.

@@

\section*{Reviewer \#2} 
\subsection*{Comment R2.1}
\begin{quote}
``I appreciate the authors' responses to my previous comments. However, despite the revisions, significant weaknesses remain, rendering in my opinion this study primarily a methodological contribution that adds little value to our understanding of interference effects. Additionally, the authors' strong claims often come through as overstatements that could result in serious misconceptions.
My overall assessment is as follows.

1. Interpretability of the SPR results: The results from the self-paced reading (SPR) task remain unclear. The only observable effect—the semantic interference—emerges well before the critical region of interest (specifically, at the distractor region) and persists without change beyond the critical region. As a result, any effect at the critical region is uninterpretable.

The authors acknowledge the interpretability issue with the SPR results in their response to my comment, and they now include the following statement: ``Given the pre-critical reading time differences, effects in the later regions, i.e., the critical and spill-over regions, cannot be attributed clearly to the stimuli in these regions and sentence processing mechanisms associated with them. Therefore, we refrain from further discussing effects occurring in the later regions." \textbf{However, this comes after more than 10 full pages (pp. 25-37) of discussion on the reaction times (RTs) in the critical and pre-critical regions, which would lead any reader to believe these findings are meaningful. If, as the authors now realize, the effect emerges much earlier, this entire discussion is misleading: the SPR results after the distractor region are simply uninterpretable.}''
\end{quote}

\subsubsection*{Response to R2.1}
We agree with the reviewer that the presentation and discussion of the SPR results was misleading and have revised it accordingly. We have revised Figure 3 to remove the focus on the critical and surrounding regions (see below). At the beginning of the Results section, we now anticipate our interpretation of the results and state that we believe that the distractors led to encoding interference and that the long-lasting effect caused by that confounds the results at later effects. For comparability with previous findings, we then briefly report the later effects. We have revised the Results section in the following way, starting on page \pageref{revised_SPR}:

@@

\begin{quote}
    \Paste{revised_SPR}
\end{quote}

We do not share the reviewer's interpretation that the effects in the later regions are not meaningful. We use the pre-critical reading times as a proxy to investigate the effect of encoding interference caused by the distractor. To emphasize this we have renamed the former section ``Pre-critical effects" to ``Reading times difference prior to retrieval". We have revised the beginning of the Discussion on page \pageref{prior_to_retrieval} accordingly:

@@

\subsubsection*{Response to R2.2}
The focus of the present study was to investigate retrieval interference due to syntactic and semantic similarity. The finding that the reading times were primarily affected by encoding interference was unexpected (but see \citeauthor{mertzen}'s (\citeyear{mertzen}) discussion). Given that the present manuscript contains two large-sample experiments and computational modeling (and its substantial length), we think that it is outside the scope of this manuscript to provide an in-depth discussion of a mechanism of encoding interference. To address the reviewer's concern that encoding interference would not cause such long-lasting effects as found in our SPR data, we have added the following text to the manuscript about a recent joint account of encoding and retrieval interference:

\textcolor{red}{Write something about Himanshu's work in relation to this}

\subsection*{Comment R2.3}

\begin{quote}
``2. Lack of syntactic interference evidence: No evidence for syntactic interference was found in either task, contradicting two previous studies that tested similar structures (Van Dyke 2007; Mertzen et al. 2023). \textbf{In my previous review, I noted that the syntactic manipulation used was not appropriate and thus unlikely to induce interference effects. Although the authors concede that this might be true, they did not address the logical implications of this acknowledgment, as evidenced by their repeated strong assertions}, such as the one in the abstract: ``Surprisingly, in both experiments, Bayes factor analyses showed evidence against interference due to syntactic cues".
If the syntactic manipulation is irrelevant, then repeatedly concluding that syntactic interference is absent is not only an overstatement but also misleading. All this study allows us to conclude is that when cues that are irrelevant to syntax are manipulated (like [+being a subject of whichever clause at whichever level of embedding], underpowered studies (Van Dyke 2007; Mertzen et al. 2023) may falsely report interference effects. While this is a valid conclusion and a sound methodological critique, it does not significantly advance our theoretical understanding of syntactic interference. Specifically, it fails to demonstrate that when relevant syntactic cues are manipulated, syntactic interference is not attested. Yet, this is the conclusion the authors consistently imply throughout their manuscript, leading to significant misconceptions. As I mentioned in my previous review, beyond the two studies the authors are building on, numerous studies in the literature show that hierarchically intervening elements do generate strong syntactic interference.
It is a fine enterprise to correct inaccurate claims from previously underpowered studies that used irrelevant manipulations; however, this should not not suggest that the findings presented here have a broader significance than they do.''
\end{quote}
\subsubsection*{Response to R2.3}
We agree with the reviewer that our present work does not rule out syntactic interference in general, but the present study using the present design did not induce considerable syntactic interference. We have made the following edits in the manuscript to emphasize this point:

@@

\textcolor{red}{revise text and paste here}

\textcolor{red}{SV email: i think we didn't write clearly enough about the fact that syntactic interference *in this design* is expected given the previous work. we did say it but needs to be clearer. also, when we say we find evidence against syn int, we should always add, in this design. and then we should say that this does not mean that there is no syntactic manipulation that would cause no interference, there could be. i think that point is valid. what i found strange in her comment was that she says both that van dyke and mertzen found syn int, but also that syn int is never going to show up in this design because it is not the right manipulation.  maybe her point is that we only show that the van dyke and mertzen result was Type M error (i.e., there is no real effect in this design).}

\subsection*{Comment R2.4}
\begin{quote}
``3. Misalignment of SPR and ERP results: The results from the self-paced reading (SPR) and event-related potential (ERP) methods do not align, with the only observed effect—semantic interference—pointing in opposite directions: inhibitory in SPR but facilitatory in ERP. Specifically, the authors report a reversed effect in the P600 component, showing a reduced P600 in conditions of higher semantic interference. This contradicts typical predictions, where the P600 is usually greater under high interference. The authors interpret this as a facilitatory interference effect (i.e., easier processing with high semantic interference). However, if this were the case, \textbf{we would also expect faster reaction times (RTs) in high semantic interference conditions, while the opposite is observed in the SPR data}.''
\end{quote}

\subsubsection*{Response to R2.4}
 \textcolor{red}{write response: no facilitation in SPR data because 1. distractor effect overshadows everything 2. mean effect in RTs cannot show inhibition and facilitation at once, inhibition was stronger}

\subsection*{Comment R2.5}
\begin{quote}
``4. Conflicting directions in ERP components: Within the ERP results, the different components also point in opposite directions. The P600 component shows a reversed effect, with a reduced P600 in conditions of high semantic interference. However, the N400 component follows the expected pattern, with stronger activation under high semantic interference. It remains unclear how the directionality of these two can be reconciled within the same theoretical framework.''
\end{quote}

\subsubsection*{Response to R2.5}

  \textcolor{red}{write response: Fleshed out more how inhibition and facilitation could both occur and manifest in ERPs}

\begin{quote}
    \Paste{inhibition_facilitation}
\end{quote}

\subsection*{Comment R2.6}
\begin{quote}
``5. Unclear conflicting mechanisms: The authors attribute the SPR results to encoding interference and the ERP results to retrieval interference, leading to a contradictory conclusion. I won't delve further into this point, as the issues raised earlier already highlight more fundamental problems.''
\end{quote}
\subsubsection*{Response to R2.6}
\textcolor{red}{Not sure whether this comment requires action. Maybe 1-2 sentences about Himanshu's work.}

\subsection*{Comment R2.7}
\begin{quote}
``Additional points:

@@

1. Original version: ``By contrast, the syntactic interference effect had only anecdotal evidence under the two narrow priors (Normal-(0, 0.1): BF10 = 2.5, Normal-(0, 0.5): BF10 = 1.2). There was evidence against syntactic interference under wider priors (Normal-(0, 1): BF10 = 0.66, Normal-(0, 5): BF10 = 0.14). What this means is that, in our data, only if we assume a priori that the effect size is relatively small, there is very weak evidence in favor of syntactic interference."

Revised version: ``By contrast, Bayes factors provided either no evidence for or even evidence against syntactic interference in both spatiotemporal windows (BF10 $<$ 1.2). Similarly, Bayes factors provided either no evidence for or even evidence against the interaction in both spatio-temporal windows (BF10 $<$ 1.2)."

\textbf{Did the authors change the analyses? Why what was reported as anecdotal evidence under narrow priors is now reported as no evidence or evidence against?}''
\end{quote}
\subsubsection*{Response to R2.7}
Yes, the priors were changed in the 1st revision (directional priors in the original version and symmetrical priors in the revised version, see also our Response to E.3). This resulted in small numerical changes of the results, e.g., BF10 = 2.5 in the original analysis decreased to BF10 = 1.2 in the revised analysis. Furthermore in the original version of the manuscript, we were inconsistent in our description of small BFs, i.e., BF10 = 1.2. Since the first revision and in line with \cite{jeffreys1998theory} and \cite{lee2014bayesian}, we now consistently interpret such small BFs as providing no evidence either way.

\subsection*{Comment R2.8}
\begin{quote}
``2. As a side note, on p. 58, the authors conclude that ``the parser searches for a subject that is within the same clause but uses the animacy cue without reference to the clause in which a noun appears." Does this mechanism truly seem plausible to the authors? What would this imply in practice? \textbf{Does it suggest that semantic and syntactic information are entirely encapsulated from each other?} If we were to set aside the issues I raised earlier, this would be one of the paper's central conclusions, yet the authors fail to provide a credible mechanism to support it.''
\end{quote}
\subsubsection*{Response to R2.8}
\textcolor{red}{SV email: the syn / sem comment is also interesting because there are some theories that only assume semantic cue association--there isn't even any syntax. some discussion of the theoretically possible connections between syn and sem processing would help. e.g., one extreme is that syn and sem operate in tandem (e.g., categorial grammars), others believe that sem processing strictly follows syntactic structure building (classical montagovian stuff), then we have the sem only models (e.g., sentence gestalt model). there is no consensus on what the relationship is between syn and sem, so these could operate independently. local coherence is also an example--if we assume syntactic processing and no independent semantic processing, local coherence would not arise (depending on how one assumes that syntactic processing works).}

\subsection*{Comment R2.9}
\begin{quote}
``3. How do the fillers differ from the experimental items? In the example the authors reported there is retrieval at place again:

@@

Experimental: The neighbor believed that the widower, who had told her that the loss was awful, regularly drank in the evenings to forget.\\
Filler: The carpet maker, who came to the workshop early, repaired the especially beautiful old carpet while listening to the news.

The authors state, ``The fillers were less syntactically complex than the experimental items but had at least one embedded clause and generally provided more variety." \textbf{In what way did they provide more variety? In which respects? Both fillers and test items involve retrieval; what is the role of the fillers in this context?}''
\end{quote}
\subsubsection*{Response to R2.9}
The fillers included different types of subordinate clauses and (more) modifiers at different positions within the sentence. By deviating from the rigid structure of the critical items, they provided more variety in regard to the employed sentence structures. To avoid creating an obvious juxtaposition of very complex sentences (critical items) and very simple sentences (fillers), the fillers included long-distance dependencies and at least one subordinate clause. We have revised the description of the fillers in the manuscript on page \pageref{fillers} in the following way:

\begin{quote}
    \Paste{fillers}
\end{quote}

\subsection*{Comment R2.10}
\begin{quote}
``4. There are numerous repetitions throughout the paper. A prime example is the paragraph on ``hyp testing using BF" (p. 45), which is repeated almost verbatim in the ``Discussion" section (p. 47). This redundancy occurs several times, making the paper excessively and unnecessarily long. The authors should streamline the content to enhance clarity and conciseness.''
\end{quote}

\subsubsection*{Response to R2.10}
\textcolor{red}{remove repetitions}

@@

\subsection*{Comment R3.3}
\begin{quote}
``Finally, and in a similar vein, I raised questions about whether there might be other reasons for the selective interference effects found in this study. Maybe a different mechanism could capture the special status of the subject-of-the-same-clause, or maybe some specific property of the current experimental materials could be responsible for the lack of interference from other subjects. The authors seem skeptical of the first of these, and they seem more sympathetic to the second. \textbf{This is all reasonable enough, and their text conveys some caution. The abstract, on the other hand, is rather more confident.} I am sympathetic to the authors' preferred conclusion. But I am less confident than they are about how well justified it is based on the current findings. This has nothing to do with the quantitative wizardry that the authors display in the paper. It's all about the issue of generalizing from a single sentence configuration (that participants saw in 60\% of trials in the study).''
\end{quote}

\subsubsection*{Response to R3.3}
We have revised the abstract to emphasize that our findings only apply to the present design. We paste the revised abstract here:

### 2024-11-11 paper/revision2/elsarticle-template-harv.tex
Add files via upload

\begin{abstract}
%% Text of abstract
\Copy{abstract}{
Cue-based retrieval accounts of sentence processing postulate that at a verb, retrieval cues are generated to complete a dependency with the verb's argument(s); for example, the dependency between the subject and the verb must be completed to interpret the sentence. If these retrieval cues match with not only the features on the noun that is the retrieval target but also with those on other nouns in the sentence, then processing difficulty arises at the verb. This difficulty in identifying the correct dependent is called similarity-based interference. 
We present large-sample self-paced reading (N = 774) and event-related potentials experiments (N = 103) using a well-established design \citep{vandyke07,mertzen}, to investigate interference due to syntactic and semantic cues in German. \textcolor{blue}{In this design, the syntactic cue \{+subject\} and the semantic cue \{+animate\} are manipulated.}
A Bayes factor analysis showed that both the self-paced reading times as well as event-related potentials provided clear evidence for a semantic interference effect. In the self-paced reading experiment, \textcolor{blue}{semantic interference due to the semantic cue \{+animate\}} caused a reading time slow-down starting at the distractor. 
The event-related potentials showed a more negative N400 amplitude at the critical verb, i.e., the retrieval site, for high vs.\ low semantic interference.
Surprisingly, in both experiments, Bayes factor analyses showed evidence against interference due to \textcolor{blue}{the syntactic cue \{+subject\}}. This finding contradicts the predictions of the standard implementations of cue-based retrieval theory, which (implicitly) assume that both syntactic and semantic cues play an equal role in retrieval. We show through computational modeling that cue-based retrieval also shows no syntactic interference \textcolor{blue}{in the present design} if the parser is assumed to keep track of which clause the subject occurs in. Thus, if syntactic retrieval cues include hierarchical syntactic information (is the noun in the same clause as the verb?), the predictions of the cue-based retrieval model would be consistent with the observed patterns in our data.

@@

Previous work has shown clear evidence for this kind of retrieval interference \citep{vandyke07, vandyke_mcelree06, vandyke_mcelree2011, nicenboim,  vandyke_lewis03,jaeger_etal_2017}. However, two important questions remain unanswered. First, as discussed in \cite{jaeger_etal_2017}, almost all the published studies on interference effects are severely underpowered. Low power has the consequence that published estimates will tend to be overestimates or misestimates of the effect size \citep{vasishth2018_signficancefilter}. Over-/misestimates have important implications for theory evaluation. One example is discussed in \cite{vasishth2018_signficancefilter}, where a key prediction of surprisal theory \citep{levy&keller_2013} could not be validated when a larger-sample study was conducted. By contrast, the original small-sample study \citep{levy&keller_2013} showed suggestive evidence consistent with the theory. Another example is the recent heated debate \citep{nieuwland_etal_2018} on prediction in sentence comprehension, which was largely focused on the results of low-powered studies, leading to potentially misleading conclusions \citep{nicenboim_etal_2020}.
Second, although it has been established \citep[e.g.,][]{vandyke07,mertzen} that syntactic and semantic cues are involved in sentence processing, it is not obvious which cues are relevant for a particular retrieval event. For example, \cite{dillon2013} argued, following \cite{Sturt2003}, that the gender cue was either downweighted or not used at all when an antecedent-reflexive dependency needs to be built. Instead, the antecedent of a reflexive could be identified using the syntactic principle A of the binding theory, which correctly identifies the target for retrieval.  Using computational modeling, \cite{dillon2013} suggested that antecedent-reflexive dependencies were immune to interference, and that the reason for this is that principle A of the binding theory is the primary driver of the cue-based retrieval process \citep[cf. ][]{jaeger_etal_2020,yadav2021individual}. This was a surprising discovery because the default assumption in cue-based retrieval accounts was that all cues were equally weighted, regardless of the dependency type; this had the consequence that retrieval interference was expected whenever a cue-overload configuration was present in a sentence. More broadly, an important insight in the work of \cite{dillon2013}, among others, is that  the precise details of which cues are used during retrieval can affect processing difficulty in possibly surprising ways. We build on this idea in the present paper.

@@

\subsubsection{Procedure}
Participants completed a moving window self-paced reading task \citep{just_etal_1982} on the PCIbex Farm \citep{pcibex}. They were instructed to read for comprehension at a comfortable pace. All sentences were displayed word by word. Masked words were presented as a line indicating its length. Unmasked words were presented in 18 pt Courier font. The space bar on the keyboard was used to unmask words. The length of the sentences required line breaks. These were hard-coded so that in each sentence a line break appeared i) between the subject and the relative clause and ii) two words after the critical verb (see \ref{ex:linebreak}, critical word in bold, | indicate line breaks in the moving window display). In a third of the trials, the sentence was followed by a yes/no comprehension question. The participants received no feedback on their performance.

@@

\begin{exe}  
\ex \label{ex:linebreak}
    \gll Die Nachbarin glaubte, dass der Witwer, | der erzählt hatte, dass der Verlust schrecklich war, regelmäßig abends \textbf{trank}, um zu | vergessen. \\ 
    The neighbor believed that the widower | who told had that the loss awful was regularly in.the.evening drank in.order to | forget\\
    \trans  `The neighbor believed that the widower, who had told her that the loss was awful, regularly drank in the evenings to forget.'\\  
\end{exe}

{In Experiment 1a, participants were re-invited for session 2 after they completed session 1. The experimental sessions were separated by 1 - 20 days. The procedure of the sessions was identical. In session 1, the experimental items 1-60 were presented interspersed with 40 filler sentences. In session 2, the experimental items 61-120 were presented with another set of 40 filler sentences. The fillers were less syntactically complex than the experimental items but had at least one embedded clause and generally provided more variety (e.g., ``Der Teppichmacher, der früh in die Werkstatt gekommen war, reparierte den besonders schönen, alten Teppich während er die Nachrichten hörte'', `The carpet maker who came to the workshop early repaired the especially beautiful old carpet while listening to the news.'; ``Der Jugendliche war genervt, weil seine Freundin, die nicht studieren wollte, oft die Schule schwänzte, um Computer zu spielen.'', `The teenager was annoyed because his girlfriend who did not want to go to University  skipped school frequently to play video games.').}\label{fillers1} Each session lasted approximately 25 minutes.

@@

Self-paced reading times were analyzed with Bayesian linear mixed models in R \citep{r} with log-normal likelihood, using the brms package \citep{brms}. The critical word for analyses was the verb \textit{trank} `drank' in (\ref{ex:materials}), constituting the retrieval site. Because previous work found effects in the pre-critical and post-critical region, reading times of the pre-critical word (\textit{regelmäßig} `regularly' in \ref{ex:materials}) and the spill-over region (\textit{um} `in.order.to' in \ref{ex:materials}) were analyzed. 
The models included fixed effects for syntactic interference (high +0.5, low -0.5), semantic interference (high +0.5, low -0.5) and their interaction. In addition to the predictors of interest, we included trial id to account for potential effects of fatigue or adaptation to the task. The trial id of the 100 trials per session was rescaled to span from 0 to 1. This was done to bring all fixed effects to a comparable scale which decreases the run time of the models. The models were run with full random effects, i.e., varying intercepts and slopes of all fixed effects and their interactions by participants and items.

@@

\Copy{trim}{{\textcolor{blue}{Reading times below 150 ms and above 3000 ms were excluded before analysis (and visualization) because these overly short or long reaction times were likely caused by inattentively performing the task. The procedure excluded 4.9\,\% of the data.}}}\label{trim} Reading times across the whole sentence are shown in Figure \ref{fig:whole_sentence}, separately for high \textcolor{blue}{(panels A and C)} and low \textcolor{blue}{(panels B and D)} syntactic interference because these conditions had partially different sentence structures.

\begin{sidewaysfigure}[h]
    \caption{Self-paced reading times with 95\% confidence intervals. Panel A and B show the pooled reading times across the whole sentence; separately for high (A) and low (B) syntactic interference due to differing sentence structure. \textcolor{blue}{Panel C and D show the reading times of the sentence focusing on the regions between the distractor and critical verb for each session of Experiments 1a and 1b; again separately for high (C) and low (D) syntactic interference.}}
    \label{fig:whole_sentence}
    \centering
    \includegraphics[width=\textwidth]{Pandora_all_wholesentence_pooled_exp.jpg}
\end{sidewaysfigure}
\clearpage

@@

\textcolor{blue}{Crucially, the reading times difference starting at the distractor} persisted in the following regions. The distractor and the immediately following regions differ between conditions, \textcolor{blue}{therefore we did not carry out any statistical tests for these observed reading time differences. The reading time difference starting at the distractor and persisting almost until the end of the sentence is problematic for the analysis and interpretation of reading times in the later regions, i.e., the critical region, because any effects that might be present there cannot be attributed clearly to the processing of that specific region. However, for comparability with previous work, we briefly report analyses of the reading times of the later regions, i.e., pre-critical, critical and spill-over region, below.}

@@

\subsection{Discussion}
\Copy{SPR_discussion_revised}{We presented the to-date largest-sample self-paced reading study that \textcolor{blue}{aimed to investigate} the use of syntactic and semantic features during subject-verb dependency formation. \textcolor{blue}{However, reading times started to differ before this critical dependency could be formed. Starting at the distractor which intervened between subject and verb, high semantic interference conditions (animate distractors) were read slower than low semantic interference conditions (inanimate distractors). Because the distractor differed between conditions, we did not analyze distractor reading times. Instead, we treat the pre-critical region which was identical across conditions and preceded the critical verb as a proxy to investigate the difference starting at the distractor. Bayes factors provided extremely strong evidence for a semantic interference effect in the pre-critical region. We refrain from interpreting effects which were observed in later regions, i.e., the critical verb and spill-over region, because it is unclear whether they were caused by the difference starting at the distractor or other processes. The reading times differences at the distractor and the pre-critical region are discussed below.}}

@@

To investigate whether this plausibility difference is driving the \textcolor{blue}{reading time difference between the semantic interference conditions}, we ran a Bayesian linear mixed model with log-normal likelihood on the reading times of the pre-critical region. This model included fixed effects for semantic interference (high +0.5, low -0.5), trial id (recoded to span from 0 to 1) and the centered plausibility ratings. We used the same priors as presented in Table \ref{tab:spr_priors}. The model was run with full random effects, i.e., varying intercepts and slopes of all fixed effects and their interactions by participants and items. Figure \ref{fig:posteriors_plausibility} presents the posteriors from the model with the Normal(0, 0.05) prior on the slopes. Although plausibility influenced reading times (higher plausibility led to faster reading times, CrI [-30, -6] ms), there was an independent effect of semantic interference (CrI [2, 24] ms). The Bayes factors provided evidence for semantic interference in this analysis under all priors except the most diffuse one (Normal(0, 0.01): BF$_{10}$ = 4.2,
Normal(0, 0.05): BF$_{10}$ = 2.4,
Normal(0, 0.1): BF$_{10}$ = 1.3). The interaction of semantic interference and plausibility was centered around zero. In conclusion, the slowdown in the reading times results for high vs.\ low semantic interference \textcolor{blue}{starting at the distractor and lasting into later regions} was partially caused by reduced plausibility of the high semantic interference conditions. However, the plausibility difference was not the only driver of the reading times slowdown, i.e., \textcolor{blue}{high} semantic interference led to slower reading times independent of plausibility.

>>>>

\begin{abstract}
%% Text of abstract
\Copy{abstract}{
Cue-based retrieval accounts of sentence processing postulate that at a verb, retrieval cues are generated to complete a dependency with the verb's argument(s); for example, the dependency between the subject and the verb must be completed to interpret the sentence. If these retrieval cues match with not only the features on the noun that is the retrieval target but also with those on other nouns in the sentence, then processing difficulty arises at the verb. This difficulty in identifying the correct dependent is called similarity-based interference. 
We present large-sample self-paced reading (N = 774) and event-related potentials experiments (N = 103) using a well-established design \citep{vandyke07,mertzen}, to investigate interference due to syntactic and semantic cues in German. \textcolor{blue}{In this design, the syntactic cue \{+subject\} and the semantic cue \{+animate\} are manipulated.}
A Bayes factor analysis showed that both the self-paced reading times as well as event-related potentials provided clear evidence for a semantic interference effect. In the self-paced reading experiment, \textcolor{blue}{semantic interference due to the semantic cue \{+animate\}} caused a reading time slow-down starting at the distractor. 
The event-related potentials showed a more negative N400 amplitude at the critical verb, i.e., the retrieval site, for high vs.\ low semantic interference.
Surprisingly, in both experiments, Bayes factor analyses showed evidence against interference due to \textcolor{blue}{the syntactic cue \{+ grammatical subject\} in this design}. This finding contradicts the predictions of the standard implementations of cue-based retrieval theory, which (implicitly) assume that both syntactic and semantic cues play an equal role in retrieval. We show through computational modeling that cue-based retrieval also shows no syntactic interference \textcolor{blue}{in the present design} if the parser is assumed to keep track of which clause the subject occurs in. Thus, if syntactic retrieval cues include hierarchical syntactic information (is the noun in the same clause as the verb?), the predictions of the cue-based retrieval model would be consistent with the observed patterns in our data.

@@

Previous work has shown clear evidence for this kind of retrieval interference \citep{vandyke07, vandyke_mcelree06, vandyke_mcelree2011, nicenboim,  vandyke_lewis03,jaeger_etal_2017}. However, two important questions remain unanswered. First, as discussed in \cite{jaeger_etal_2017}, almost all the published studies on interference effects are severely underpowered. Low power has the consequence that published estimates will tend to be overestimates or misestimates of the effect size \citep{vasishth2018_signficancefilter}. Over-/misestimates have important implications for theory evaluation. One example is discussed in \cite{vasishth2018_signficancefilter}, where a key prediction of surprisal theory \citep{levy&keller_2013} could not be validated when a larger-sample study was conducted. By contrast, the original small-sample study \citep{levy&keller_2013} showed suggestive evidence consistent with the theory. Another example is the recent heated debate \citep{nieuwland_etal_2018} on prediction in sentence comprehension, which was largely focused on the results of low-powered studies, leading to potentially misleading conclusions \citep{nicenboim_etal_2020}.
Second, although it has been established \citep[e.g.,][]{vandyke07,mertzen} that syntactic and semantic cues are involved in sentence processing, it is not obvious which cues are relevant for a particular retrieval event. For example, \cite{dillon2013} argued, following \cite{Sturt2003}, that the gender cue was either downweighted or not used at all when an antecedent-reflexive dependency needs to be build. Instead, the antecedent of a reflexive could be identified using the syntactic principle A of the binding theory, which correctly identifies the target for retrieval.  Using computational modeling, \cite{dillon2013} suggested that antecedent-reflexive dependencies were immune to interference, and that the reason for this is that principle A of the binding theory is the primary driver of the cue-based retrieval process \citep[cf. ][]{jaeger_etal_2020,yadav2021individual}. This was a surprising discovery because the default assumption in cue-based retrieval accounts was that all cues were equally weighted, regardless of the dependency type; this had the consequence that retrieval interference was expected whenever a cue-overload configuration was present in a sentence. More broadly, an important insight in the work of \cite{dillon2013}, among others, is that  the precise details of which cues are used during retrieval can affect processing difficulty in possibly surprising ways. We build on this idea in the present paper.

@@

\subsubsection{Procedure}
Participants completed a moving window self-paced reading task \citep{just_etal_1982} on the PCIbex Farm \citep{pcibex}. They were instructed to read for comprehension at a comfortable pace. All sentences were displayed word by word. Masked words were presented as a line indicating its length. Unmasked words were presented in 18 pt Courier font. The space bar on the keyboard was used to unmask words. The length of the sentences required line breaks. These were hard-coded so that in each sentence a line break appeared i) between the subject and the relative clause and ii) two words after the critical verb (see \ref{ex:linebreak}, critical word in bold, $\mid$ indicate line breaks in the moving window display). In a third of the trials, the sentence was followed by a yes/no comprehension question. The participants received no feedback on their performance.

@@

\begin{exe}  
\ex \label{ex:linebreak}
    \gll Die Nachbarin glaubte, dass der Witwer, $\mid$ der erzählt hatte, dass der Verlust schrecklich war, regelmäßig abends \textbf{trank}, um zu $\mid$  vergessen. \\ 
    The neighbor believed that the widower $\mid$  who told had that the loss awful was regularly in.the.evening drank in.order to $\mid$  forget\\
    \trans  `The neighbor believed that the widower, who had told her that the loss was awful, regularly drank in the evenings to forget.'\\  
\end{exe}

In Experiment 1a, participants were re-invited for session 2 after they completed session 1. The experimental sessions were separated by 1 - 20 days. The procedure of the sessions was identical. In session 1, the experimental items 1-60 were presented interspersed with 40 filler sentences. In session 2, the experimental items 61-120 were presented with another set of 40 filler sentences. {\Copy{fillers}{\textcolor{blue}{The fillers were included to mask the critical manipulation. They} were less syntactically complex than the experimental items but had at least one embedded clause and generally provided more variety\textcolor{blue}{, e.g., by including modifiers and different types of subordinate clauses} (e.g., ``Der Teppichmacher, der früh in die Werkstatt gekommen war, reparierte den besonders schönen, alten Teppich während er die Nachrichten hörte'', `The carpet maker who came to the workshop early repaired the especially beautiful old carpet while listening to the news.'; ``Der Jugendliche war genervt, weil seine Freundin, die nicht studieren wollte, oft die Schule schwänzte, um Computer zu spielen.'', `The teenager was annoyed because his girlfriend who did not want to go to University  skipped school frequently to play video games.').}}\label{fillers} Each session lasted approximately 25 minutes.

@@

\Copy{trim}{{\textcolor{blue}{Self-paced reading times below 150 ms and above 3000 ms were excluded before analysis because these overly short or long reaction times were likely caused by inattentively performing the task. The procedure excluded 4.9\,\% of the data.}}}\label{trim} Self-paced reading times were analyzed with Bayesian linear mixed models in R \citep{r} with log-normal likelihood, using the brms package \citep{brms}. The critical word for analyses was the verb \textit{trank} `drank' in (\ref{ex:materials}), constituting the retrieval site. Because previous work found effects in the pre-critical and post-critical region, reading times of the pre-critical word (\textit{regelmäßig} `regularly' in \ref{ex:materials}) and the spill-over region (\textit{um} `in.order.to' in \ref{ex:materials}) were analyzed. 
The models included fixed effects for syntactic interference (high +0.5, low -0.5), semantic interference (high +0.5, low -0.5) and their interaction. In addition to the predictors of interest, we included trial id to account for potential effects of fatigue or adaptation to the task. The trial id of the 100 trials per session was rescaled to span from 0 to 1. This was done to bring all fixed effects to a comparable scale which decreases the run time of the models. The models were run with full random effects, i.e., varying intercepts and slopes of all fixed effects and their interactions by participants and items.

@@

Reading times across the whole sentence are shown in Figure \ref{fig:whole_sentence}, separately for high \textcolor{blue}{(panels A and C)} and low \textcolor{blue}{(panels B and D)} syntactic interference because these conditions had partially different sentence structures.

\begin{sidewaysfigure}[h]
    \caption{Self-paced reading times with 95\% confidence intervals. Panel A and B show the pooled reading times across the whole sentence; separately for high (A) and low (B) syntactic interference due to differing sentence structure. \textcolor{blue}{Panel C and D show the reading times of the sentence focusing on the regions between the distractor and critical verb for each session of Experiment 1a and 1b; again separately for high (C) and low (D) syntactic interference.}}
    \label{fig:whole_sentence}
    \centering
    \includegraphics[width=\textwidth]{Pandora_all_wholesentence_pooled_exp.jpg}
\end{sidewaysfigure}
\clearpage

@@

\textcolor{blue}{Crucially, the reading times difference starting at the distractor} persisted in the following regions. The distractor and the immediately following regions differed between conditions, \textcolor{blue}{therefore we did not analyze their reading times. The reading times difference starting at the distractor and persisting almost until the end of the sentence is problematic for the analysis and interpretation of reading times in the later regions, i.e., the critical region. Any effects that might be present there cannot be attributed clearly to the processing of the respective region. However, for comparability with previous work, we briefly report analyses of the reading times of the later regions, i.e., the pre-critical, critical and spill-over region.}

@@

\subsection{Discussion}
\Copy{SPR_discussion_revised}{We presented the to-date largest-sample self-paced reading study that \textcolor{blue}{aimed to investigate} the use of syntactic and semantic features during subject-verb dependency formation. \textcolor{blue}{However, reading times started to differ before this critical dependency could be formed. Starting at the distractor which intervened between subject and verb, high semantic interference conditions (animate distractors) were read slower than low semantic interference conditions (inanimate distractors). Because the distractor differed between conditions, we did not analyze distractor reading times. Instead, we treat the pre-critical region which was identical across conditions and preceded the critical verb as a proxy to investigate the difference starting at the distractor. Bayes factors provided extremely strong evidence for a semantic interference effect in the pre-critical region. In contrast, there was evidence against syntactic interference and an interaction in that region. We refrain from interpreting effects which were observed in later regions, i.e., the critical verb and spill-over region, because it is unclear whether they were caused by the difference starting at the distractor or other processes. The reading times differences at the distractor and the pre-critical region are discussed below.}}

@@

To investigate whether this plausibility difference was driving the \textcolor{blue}{reading time difference between the semantic interference conditions}, we ran a Bayesian linear mixed model with log-normal likelihood on the reading times of the pre-critical region. This model included fixed effects for semantic interference (high +0.5, low -0.5), trial id (recoded to span from 0 to 1) and the centered plausibility ratings. We used the same priors as presented in Table \ref{tab:spr_priors}. The model was run with full random effects, i.e., varying intercepts and slopes of all fixed effects and their interactions by participants and items. Figure \ref{fig:posteriors_plausibility} presents the posteriors from the model with the Normal(0, 0.05) prior on the slopes. Although plausibility influenced reading times (higher plausibility led to faster reading times, CrI [-30, -6] ms), there was an independent effect of semantic interference (CrI [2, 24] ms). The Bayes factors provided evidence for semantic interference in this analysis under all priors except the most diffuse one (Normal(0, 0.01): BF$_{10}$ = 4.2,
Normal(0, 0.05): BF$_{10}$ = 2.4,
Normal(0, 0.1): BF$_{10}$ = 1.3). The interaction of semantic interference and plausibility was centered around zero. In conclusion, the slowdown in the reading times results for high vs.\ low semantic interference \textcolor{blue}{starting at the distractor and lasting into later regions} was partially caused by reduced plausibility of the high semantic interference conditions. However, the plausibility difference was not the only driver of the reading times slowdown, i.e., \textcolor{blue}{high} semantic interference led to slower reading times independent of plausibility.
